\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{url}
\usepackage{hyperref}


\begin{document}

\title{Classification of Selected Philippine Bird Species Using Acoustic Features and Classical Machine Learning}
\titlerunning{Philippine Bird Species Classification}

\author{Emmanuel Albeza \and Yvonne Lin \and Christalie Pategara}
\institute{University of the Philippines Visayas\\
\email{eaalbeza@up.edu.ph, yalin@up.edu.ph, chpategara@up.edu.ph}}

\maketitle

\begin{abstract}
This study proposes a classical machine learning approach for the classification of selected Philippine bird species from audio recordings. The study will use recordings from the Xeno-canto bird sound repository, extract acoustic features such as Mel-frequency cepstral coefficients (MFCCs) and spectral descriptors, and compare traditional classifiers including Logistic Regression, Support Vector Machine, Random Forest, and XGBoost. The project aims to assess the feasibility of bird-call recognition for biodiversity monitoring and assisted species identification.
\keywords{bird sound classification \and acoustic features \and machine learning \and Philippine birds}
\end{abstract}

\section{Description of the Project}

This study proposes a classical machine learning model for classifying selected Philippine bird species from audio recordings. Bird vocalizations are suitable for biodiversity monitoring because passive acoustic monitoring provides an effective and non-invasive means of detecting vocally active species and can support broader temporal and spatial coverage than observer-based surveys in some settings \cite{szymanski2021,biffi2024,mosikidi2023}. Automated bird-call classification may therefore support ecological monitoring and species identification while reducing the effort required for manual review. To address this objective, the study will use recordings from Xeno-canto, a public wildlife sound repository that provides downloadable bird recordings and metadata \cite{xenocanto}.

Bird-sound classification has practical applications in biodiversity assessment, passive acoustic monitoring, and educational tools for species recognition. Previous studies have shown that bird sounds can be classified using handcrafted acoustic features, particularly Mel-frequency cepstral coefficients (MFCCs), together with traditional machine learning models such as Support Vector Machines. Earlier work employed mel-cepstral and low-level acoustic parameters with SVM classifiers, while more recent studies have used MFCCs with ECOC-SVM and compared them with other traditional models, including Random Forest \cite{fagerlund2007,han2023}. Building on this body of work, the proposed study focuses on a selected subset of Philippine bird species and adopts classical machine learning methods.

The study addresses the difficulty of species-level identification from field recordings, which typically requires manual effort and domain expertise. A successful classifier would demonstrate the feasibility of automated acoustic recognition as a support tool for biodiversity monitoring.
\section{Proposed Method}

This study will follow a structured process comprising data collection, audio recording pre-processing, feature extraction, model training, and evaluation.

The data collection will be conducted by utilizing the Xeno-canto repository, from which metadata and audio recordings of selected Philippine bird species will be collected \cite{xenocanto}. The metadata provided by the repository, such as scientific names and song types, will be used later on to aid in the classification process. For the pre-processing, audio recordings will be standardized to a uniform sampling rate, trimmed or padded to ensure each clip has the same duration. Moreover, audio recordings that have poor or unusable audio quality will be removed during this step. 

After pre-processing, the processed audio recordings will undergo feature extraction using Mel-frequency cepstral coefficients (MFCCs), a feature extraction technique widely used in audio classification tasks by converting an audio signal into a set of numerical coefficients \cite{librosa_mfcc,fagerlund2007}. This technique was chosen because previous studies have proven that it effectively correlates features and captures the unique vocal characteristics of birds by how sound is distributed across different frequencies. In addition to MFCCs, this step will also extract descriptive signal parameters \cite{fagerlund2007}:

{
\renewcommand{\labelitemi}{\small$\bullet$}
\begin{itemize}
  \item Spectral features: These include the spectral centroid, bandwidth, and range, which describe the frequency characteristics and distribution of the sound’s spectrum.
  \item Temporal features: These include the zero-crossing rate, short-time energy, and duration, which describe the intensity of the sound wave over time.
\end{itemize}
}

For model training and selection, this study will evaluate and compare different traditional machine learning classifiers, specifically Logistic Regression, Support Vector Machine (SVM), Random Forest, and XGBoost, to determine the most effective approach for the classification of selected Philippine bird species based on the dataset collected by this study.

\section{Dataset}

The primary dataset source that will be used for this study is the Xeno-canto repository. Xeno-canto is a publicly accessible archive of wildlife recordings and associated metadata from across the world, including bird sounds. It also supports audio search and retrieval through both its website and API \cite{xenocanto}.

The dataset consists of recordings from selected bird species located in the Philippines. Rather than having a fixed final species list at the outset, the study will curate a list of species that satisfies practical requirements such as sufficient recording counts, acceptable recording quality, and manageable class balance. This strategy is intended to produce a realistic dataset for a small-scale multiclass classification task. The final subset is expected to include approximately five bird species, depending on recording availability, audio quality, and class balance.

Based on the number of recordings in the repository that can be found in the Philippines excluding the unidentified audio files, the top 5 most entries are as follows: 

1. Hypsipetes philippinus

2. Phapitreron leucotis 

3. Dicrurus balicassius 

4. Ninox philippensis 

5. Brachypteryx montana 

Since data preparation is a major component of machine learning, the study will also include dataset curation tasks such as metadata cleaning, noise screening, clip standardization, and the preparation of labeled feature tables for model training and testing.

\section{Metrics of Evaluation}

Because the task is a multiclass classification problem, model performance will be evaluated using accuracy, precision, recall, macro F1-score, and confusion matrices. Macro F1-score is particularly important because the class distribution may not be perfectly balanced. The study will use a stratified train-test split and cross-validation \cite{sklearn_f1}.

\section{Tools and Packages}

This study will be conducted in Python using the Jupyter Notebook environment for documentation and the development process, and GitHub for version control and to host the study repository. In terms of data handling and visualization, this study will utilize Pandas, Numpy, Matplotlib, and Seaborn. Furthermore, the librosa package will be the primary tool for audio processing and the extraction of MFCCs \cite{librosa_mfcc}. On the other hand, the scikit-learn library will be used to implement and compare the traditional machine learning classifiers through evaluation metrics.

\bibliographystyle{splncs04}
\bibliography{references}

\end{document}